---
title: "Boundary lines for yield gap analysis: Revisiting Fermont et al. (2009)"
author:
- Tomás Roquette Tenreiro$^a$ (PhD) and João Vasco Silva$^b$ (PhD)
- $^a$SISTAGRO - Systems Engineering in Agriculture
- $^b$CIMMYT-Zimbabwe & WUR-PPS
date: "`r format(Sys.time(), '%d-%B-%Y')`"
output:
  html_document:
    toc: yes
    toc_depth: 2
    toc_float:
      collapsed: no
      smooth_scroll: yes
    fig_caption: yes
    css: style.css
    header-includes:
    - \usepackage{setspace}
    - \sinlgespacing
  pdf_document:
    toc: yes
    toc_depth: '2'
  word_document:
    toc: yes
    toc_depth: '2'
bibliography: references.bib
csl: https://raw.githubusercontent.com/citation-style-language/styles/master/ecology-letters.csl
---

```{r, echo=FALSE}
knitr::opts_chunk$set(message = FALSE)
```

___

# **Introduction**

**Please review my text carefully and modify whatever you consider necessary**

Sustaining food production in the years ahead will require productivity gains in resource use. Specific challenges include estimating the magnitude and thus the value of yield gaps, identifying limiting factors, and implementing profitable and sustainable  strategies. Closing yield gaps in arable farming, while improving resource use  efficiency, is also expected to minimize the expansion of arable land and emissions while reducing other undesirable externalities. 

Over the last years, new methodological approaches have been evolving within the context of yield gap assessment, mostly due to advances in data engineering and programming languages. One clear example is the mathematical computation of boundary-lines over crop yield datasets to decompose yield gaps (INCLUDE SOME REFERENCES, e.g., Silva et al.). The boundary-line analysis (BLA), based on the Law of the Minimum, is a powerful concept in agronomy which can be used for quantitative decomposition of yield gaps. The BLA is considered a useful complement to more traditional data analysis techniques. From a conceptual point of view, different approaches to fit boundary lines on crop data have been proposed throughout the last century. Five different methodological families can be identified:

1.	Curves fitted “manually and by eye” (e.g., French & Schultz, 1984);
2.	The use of crop simulation modelling for benchmarking potential yields (e.g., Passioura & Angus, 2010);
3.	Empirically based methods: grouping data and use ‘splines’ (e.g., Fermont el al., 2009);
4.	Econometrics based on the use of quantile regression analysis (e.g., Grassini et al., 2015);
5.	Stochastic frontier analysis (e.g., Silva et al., 2017);

Herewith, we focused on the ‘groups and splines’ methodological family due to the following reasons:  

•	High popularity in literature (e.g., Casanova et al.);
•	Purely statistical and comprehensible approach (Fermont et al.) of plausible implementation with a R-notebook structure.
•	Not based on ’machine learning’ methods neither crop simulation modelling which would require more complex computation.
•	Easier adoption and implementation by agronomists and scientists working on crop production or research. 

The ‘group and spline’ BLA method is defined by the following general features:

•	To identify the boundary points;
•	To fit a continuous function at these points;
•	To estimate the yield gap for each production factor;
•	To identify the most limiting factor, by analogy with von Liebig’s law of the minimum, and decompose the yield gap for each field;

The yield gap decomposition can be implemented according to the method described by Wairegi et al. (2010). This method assumes responses according to von Liebig’s law of the minimum, and identifies the minimum predicted yield (Ymin) for each field as Ymin= Min(Yx1,Yx2,...,Yxn), where Yx is the corresponding BLA predicted yield for a given X-variable. The identified yield gap (IYG) is defined as the difference between the maximum attainable yield and the Ymin. The unexplained yield gap (UYG) is defined as the difference between Ymin and observed yield. The BLA serves to rank yield-constraints and to deliver a quantitative estimation of the technical issues causing yield gaps. In locations where biotic constraints are particularly important, the BLA seems to predict yields less accurately due to the higher interactions between constraints. Therefore, only the IYG can be associated with the variation of the corresponding independent variable X, which is determined by the lowest BLA-predicted yield as defined by the law of the minimum. 

This approach has been successfully used for yield gap assessment in cereal-based systems (e.g.Casanova et al., 1999; Shatar and McBratney,2004), but also for other crops such as Cassava (Fermont et al., 2009). Cassava yields in Africa are small and it remains unclear which factors most limit yields. Using a series of farm surveys and trials in Uganda and western Kenya, Fermont el al. (2009) evaluated the importance of abiotic, biotic and associated crop management constraints for cassava production. Fermont et al. (2009) started by sorting the independent variables and removing outliers. Then, BLA was used that represent the maximum yield response (the dependent variable) to the various independent variables considered. 

Here, we aim to replicate the concept of analysis conducted by Fermont et al. (2009) on the same original dataset by developing an automated analytical algorithm in R computing language. Clear boundary lines were identified in the scatter plots relating soil fertility, soil texture, pest and disease, weed management and selected rainfall variables to cassava yield. 

___

# **Load required R packages**

First, we need to load the R packages needed to run this workflow.

```{r, warning=FALSE}
# package names
packages <- c("splines", "Metrics", "dplyr", "tidyr", "knitr", "reshape2", "ggplot2", "DT")

# install packages
installed_packages <- packages %in% rownames(installed.packages())
if (any(installed_packages == FALSE)){
  install.packages(packages[!installed_packages])}

# load packages
invisible(lapply(packages, library, character.only = TRUE))
```
___

# **Crop field data**

The first step is to load the crop data to be used for yield gap analysis. The dataset contains reported observed yields, management practices, and biophysical conditions at field level (Fermont et al., 2009). 

The farm surveys and agronomic trials were carried out in arange of sites in western Kenya and central and eastern Uganda. The sites were chosen to represent a range of environments andmanagement practices in cassava-based cropping systems. In total, 13 different yield-explaining variables (x-variables) were collected for a total number of 122 fields. The following x-variables are used: 

1. "pH_soil"
2. "SOC"
3. "totN_soil"
5. "P_soil"
6. "K_soil"
7. "CBB_T0_9"
8. "CGM_T0_9"
9. "RF_tot"
10. "RF0_6M"
11. "RF_.d"
12. "clay"
13. "days_to_harvest"

```{r}
# read .csv file with data
file <- 'https://raw.githubusercontent.com/jvasco323/EiA_BLA_workflow/main/fermont_etal_final.csv'
data <- read.csv(url(file))

# list variables of interest for visualization 
str(data)
```

___

# **Data manipulation**

Here we subset all numeric independent variables by considering all points with yield values above '30 ton/ha' as outliers.

```{r}
# remove outliers
data <- subset(data, yld_t0 < 30)
```

___

# **Descriptive statistics**

Descriptive statistics of the actual yield and crop management variables used in the analysis for the 2 years of the analysis are provided below. Actual yield of Cassava across the study region was on average 13.9 t/ha. Local soils are relatively acidified (average pH < 5.8) and season rainfall was on average 1728 mm. Soil clay content and soil organic C were on average 27.2% and 10.5 mg/kg, respectively.  

```{r, echo=FALSE, warning=FALSE}
vars3 <- c('year', 'RF_tot', 'clay', 'pH_soil', 'SOC', 'days_to_harvest', 'yld_t0')
# mean
numeric_cols_mean <- data[,vars3] %>%
  group_by(year) %>%
  summarise(across(
    .cols = where(is.numeric), 
    .fns = list(Mean = mean), na.rm = TRUE, 
    .names = "{col}"))
numeric_cols_mean <- round(numeric_cols_mean, 1)
numeric_cols_mean <- t(numeric_cols_mean)
colnames(numeric_cols_mean)[2] <- 'Mean 2005'
colnames(numeric_cols_mean)[1] <- 'Mean 2004'
numeric_cols_mean = numeric_cols_mean[-1,]
Variable <- rownames(numeric_cols_mean)
rownames(numeric_cols_mean) <- NULL
numeric_cols_mean <- cbind(Variable, numeric_cols_mean)
# sd
numeric_cols_sd <- data[,vars3] %>%
  group_by(year) %>%
  summarise(across(
    .cols = where(is.numeric), 
    .fns = list(SD = sd), na.rm = TRUE, 
    .names = "{col}"))
numeric_cols_sd <- round(numeric_cols_sd, 1)
numeric_cols_sd <- t(numeric_cols_sd)
colnames(numeric_cols_sd)[2] <- 'StDev 2005'
colnames(numeric_cols_sd)[1] <- 'StDev 2004'
numeric_cols_sd = numeric_cols_sd[-1,]
Variable <- rownames(numeric_cols_sd)
rownames(numeric_cols_sd) <- NULL
numeric_cols_sd <- cbind(Variable, numeric_cols_sd)
# merge
numeric_cols = merge(numeric_cols_mean, numeric_cols_sd, by='Variable')
numeric_cols$Variable[1] <- 'Soil clay content (%)'
numeric_cols$Variable[2] <- 'Days to harvest (d)'
numeric_cols$Variable[3] <- 'Soil pH'
numeric_cols$Variable[4] <- 'Growing season rainfall (mm)'
numeric_cols$Variable[5] <- 'Soil organic C (mg/kg)'
numeric_cols$Variable[6] <- 'Cassava yield (t/ha)'
# show
knitr::kable(numeric_cols)
```

___

# **Step 1: Field selection**

This function categorizes the X-variable in 10 groups and it fits a non-linear reg. through these selected Xi and Y values (i.e., spline model). We fit a boundary line according to Fermont et al. (2009), which belongs to the methodological family of 'group + splines regressions'. This is mostly based on the use of 'Log functions' as originally proposed by Webb. (1972), Casanova et al. (1999) and Shatar & Mcbratney. (2004).

Two different approaches to identify the ymax are available in the following code chunk: 

1) ymax defined by the absolute maximum.
2) ymax defined by the mean of the highest quantile. 

The selection of the most appropriate option depends on the vertical scatter at the highest yield levels and whether the obtained BL error is acceptable or not. In our perspective, the selection of ymax based on the absolute maximum is more appropriate for denser scatters, where the maximum values approximate the main cloud of observations. The second option, which uses the highest quantile, can be a more conservative option, and a better alternative for situations when a large dispersion of individual maximum yields takes place. For those cases, the mean of the highest quantile will reduce the individual weigth of single points, which could be overestimated. It is important to note that BLA based on splines tends to underestimate ymax for most initial x-intervals (i.e., where 'y' responds largely to a variation of 'x'). In this sense, the use of the mean of the highest quantile should be avoided everytime there is enough confidence on the reported y values.  

```{r, warning=FALSE}
bl_points <- function(df, xvar, approach){
  
   # select the x variable of interest
   df <- df[,c("yld_t0", xvar)]
   
   # modify column names for generic use 
   colnames(df)[1] = "Y"
   colnames(df)[2] = "X"
   df <- subset(df, X>0, select=c(Y, X)) # why x > 0?
   
   # 'NULL' values are excluded to avoid data transformation problems and calculation failures 
   # correct NA values for both Y and X variables 
   df$Y[is.na(df$Y)] <- mean(df$Y, na.rm=T)
   df$X[is.na(df$X)] <- mean(df$X, na.rm=T)
   
   # split X variable in 10 quantiles (Fermont et al., 2009)
   x_0.1 <- subset(df, X <= quantile(X, 0.1))
   x_0.2 <- subset(df, X > quantile(X, 0.1) & X <= quantile(X, 0.2))
   x_0.3 <- subset(df, X > quantile(X, 0.2) & X <= quantile(X, 0.3))
   x_0.4 <- subset(df, X > quantile(X, 0.3) & X <= quantile(X, 0.4))
   x_0.5 <- subset(df, X > quantile(X, 0.4) & X <= quantile(X, 0.5))
   x_0.6 <- subset(df, X > quantile(X, 0.5) & X <= quantile(X, 0.6))
   x_0.7 <- subset(df, X > quantile(X, 0.6) & X <= quantile(X, 0.7))
   x_0.8 <- subset(df, X > quantile(X, 0.7) & X <= quantile(X, 0.8))
   x_0.9 <- subset(df, X > quantile(X, 0.8) & X <= quantile(X, 0.9))
   x_1.0 <- subset(df, X > quantile(X, 0.9) & X <= quantile(X, 1.0))

   # define boundary points for each quantile based on maximum value
   if(approach == 'maximum'){
    blp_0.0 <- subset(x_0.1, X == min(X)) #  briefly explain why needed
    blp_0.1 <- subset(x_0.1, Y == max(Y))
    blp_0.2 <- subset(x_0.2, Y == max(Y))
    blp_0.3 <- subset(x_0.3, Y == max(Y))
    blp_0.4 <- subset(x_0.4, Y == max(Y))
    blp_0.5 <- subset(x_0.5, Y == max(Y))
    blp_0.6 <- subset(x_0.6, Y == max(Y))
    blp_0.7 <- subset(x_0.7, Y == max(Y))
    blp_0.8 <- subset(x_0.8, Y == max(Y))
    blp_0.9 <- subset(x_0.9, Y == max(Y))
    blp_1.0 <- subset(x_1.0, Y == max(Y)) 

   # define boundary points for each quantile based on yields in given quantile
   } else if(approach == '95_quantile'){
    blp_0.0 <- subset(x_0.1, X == min(X)) 
    blp_0.1 <- subset(x_0.1, Y > quantile(Y, 0.95))
    blp_0.2 <- subset(x_0.2, Y > quantile(Y, 0.95))
    blp_0.3 <- subset(x_0.3, Y > quantile(Y, 0.95))
    blp_0.4 <- subset(x_0.4, Y > quantile(Y, 0.95))
    blp_0.5 <- subset(x_0.5, Y > quantile(Y, 0.95))
    blp_0.6 <- subset(x_0.6, Y > quantile(Y, 0.95))
    blp_0.7 <- subset(x_0.7, Y > quantile(Y, 0.95))
    blp_0.8 <- subset(x_0.8, Y > quantile(Y, 0.95))
    blp_0.9 <- subset(x_0.9, Y > quantile(Y, 0.95))
    blp_1.0 <- subset(x_1.0, Y > quantile(Y, 0.95))}

   # bind subsets
   blp_df <- rbind(blp_0.0, blp_0.1, blp_0.2, blp_0.3, blp_0.4, blp_0.5, 
                   blp_0.6, blp_0.7, blp_0.8, blp_0.9, blp_1.0)
   return(blp_df)}
```
___

# **Step 2: Boundary lines**

First, we select the variables for which BL will be fitted. 

We should avoid using variables following a discrete distributions of 'x' values. The variables 'CAD_T0_9', 'RF_tot', 'RF0_6M', 'RF_.d' and 'days_to_harvest' in Fermont et al. (2009) do not show a continuous distribution of values as these are distributed over a limitted group of X-classes. This is problematic for BLA because many observation points would be associated with the same y_predicted. The most clear case is 'CAD_T0_9' where almost all points are equal to zero. This results in an excessive concentration of points around the same y_predicted (x=0). Therefore, we decided to remove all these variables from the yield decomposition analysis.

For the selection of variables, we recommend the following two 'rules of thumb': 

1) Avoid the use of uncontinuous x-variables (i.e., discrete, partially classified, excessive skewness of x-data).

2) Avoid the use of variables for which the BL shows an excessive root mean square error. If the predictive model has a large error it shouldn't be used. The error levels should be similar among the selected boundaries and x-variables. In this specific case, we established a minimum error threshold for adoption of rmse < 3.5 ton/ha.

```{r, warning=FALSE}
# We rejected all variables not complying with the previous two rules
#variaveis <- c("pH_soil", "SOC", "totN_soil", "P_soil", "K_soil", "CBB_T0_9",
               #"CGM_T0_9", "RF_tot", "RF0_6M", "RF_.d", "clay", "days_to_harvest")
               
# We decide to move forward with only four variables               
variaveis <- c("pH_soil", "totN_soil", "P_soil", "clay")
```

Second, we create empty dataframes to bind all the new data generated in the loop below. 

```{r, warning=FALSE}
blp_new <- c()
data_new <- c()
rmse_df <- c()
```

The following loop is divided into six subsequent steps. The loop is used to create a new column (i) in the corresponding dataframes for each x-variable (i) that is selected (STEP 1). The new column (i) stores the corresponding BL points selected through the approach "maximum" (STEP 2). For each cluster of BL points, the predictive model is created (STEP 3). Then, the predictive model is used to estimate to y_max in two different data frames (STEP 4): one data frame includes only the BL points ("data.bla") and the other includes all the observation points ("data_subset"). After adding a new column (y_pred) to each dataframe, the root mean square error ("rmse") is computed on the small dataframe, wich is used to compare the maximum observed yields against the maximum predicted yields by the BL (STEP 5). Finally, all complete dataframes (each "data_subset" for each i-variable) are binded together (STEP 6). Three final tables are produced: 

1. "blp_new": only the BL points (one column with the observed yields plus n columns, each showing the predicted y_max for each i variable).
2. "data_new": all points (one column with the observed yields plus n columns, each showing the predicted y_max for each i variable).
3. "rmse_df": all the rmse (one column associated with each BL for each i-variable).

```{r, warning=FALSE}
for(i in unique(variaveis)){
  # print(i)
  
  # 1) select data
  data_subset <- data[,c("year", "Trial", "Site", "Farm.no.", "yld_t0", i)]
  colnames(data_subset)[5] = "Y"
  colnames(data_subset)[6] = "X"
  data_subset$variable <- i
  
  # 2) estimate BL points
  data.bla <- bl_points(data, i, approach="maximum")
  data.bla$variable <- i
  
  # 3) create the predictive model
  model <- glm(Y ~ ns(X, df = 2), data = data.bla)
  # print(model)
  
  # 4) predict y_max
  # boundary points only
  data.bla$y_pred <- predict(model, newdata = data.bla)
  # raw data
  data_subset$y_pred <- predict(model, newdata = data_subset)

  # 5) ERROR assessment
  rmse_value <- rmse(data.bla$Y, data.bla$y_pred)
  
  # For plotting the BLA outcomes
  plot_lims <- range(data_subset$X, na.rm=T)
  plot.grid <- seq(from=plot_lims[1], to=plot_lims[2])
  pred <- predict(model, newdata=list(X=plot.grid), se=T)
  spline.d <- as.data.frame(spline(data.bla$X, data.bla$y_pred))
  xmin <- min(data_subset$X)
  xmax <- max(data_subset$X)
  ymax <- max(data_subset$Y)
  g1 <- ggplot(data.bla, aes(x=X, y=y_pred)) + 
    geom_point(size=6, alpha=0.6, color="blue") +
    geom_point(data=data.bla, aes(x=X, y=Y), size=3, alpha=0.9, color="red") +
    geom_point(data=data_subset, aes(x=X, y=Y)) +
    coord_cartesian(xlim=c(xmin-2, xmax), ylim=c(0, ymax+5)) + 
    labs(title="BOUNDARY LINE ", subtitle="Spline method - Shatar & Mcbratney. (2004)", y="YIELD_tha",
         x="X", caption="[BL from 10 groups]") + 
    theme_bw() + 
    theme(text = element_text(size=15)) +
    geom_line(data = spline.d, aes(x = x, y = y), alpha=0.6, color="blue", size=2)
  print(g1)
  
  # 6) bind all data together 
  blp_new <- rbind(blp_new, data.bla)
  data_new <- rbind(data_new, data_subset)
  rmse_df <- rbind(rmse_df, cbind(i, rmse_value))}
```

___

# **Step 3: Yield gap analysis**

First, the data frame is reshaped to estimate the y_pred_min and the corresponding limiting factor (i.e., the corresponding x-variable with the lowest BL predicted yield for a certain observed yield).

```{r, warning=FALSE}
# reshape df
data_ygd <- dcast(data_new, year + Trial + Site + Farm.no. + Y ~ variable, value.var='y_pred') 
data_ygd$y_pred_min = apply(data_ygd[6:9], 1, min) 
data_ygd$factor_limitante = names(data_ygd[6:9])[apply(data_ygd[6:9], 1, which.min)]
```

Second, we estimate the 'identified' yield gap (IYG) and the 'unexplained' yield gap (UYG) according to Wairegi et al. (2010). 

```{r, warning=FALSE}
# decompose yg
data_ygd$IYG <- max(data_ygd$Y, na.rm=T) - data_ygd$y_pred_min
data_ygd$UYG <- data_ygd$y_pred_min - data_ygd$Y
```

Third, we plot the yield gap decomposition figure and the graphic is interpreted.

The first plot shows predicted versus observed cassava yields for the 2004 and 2005 trials. Predictions correspond to the minimum yield predicted by all the boundary lines considered. Diagonal solid lines represent the 1:1 and 2:1 lines, which indicates yield gap zones (i.e., zero to 100% gap and 100% to 200% gap). Points above the 1:1 line correspond to observed yields greater than the corresponding predictions. Those are points to which the useability of this approach fails. Those points indicate the general trend of BL for underpredicting the maximum yield on about 1/5 of the observations.

The second plot shows observed and predicted Cassava yields from the BLA. The predicted yield was the minimum prediction based on biophysical factors. Green and red vertical segments indicate the magnitude of the UYG and the IYG, respectively. The larger is the relative distance of the red line in comparison to the green one, the greater will be the fraction of total yield gap that is actually explained by the considered x-variables. Dots below the line 1:1, correpond to the same points shown above the same line in the previous plot. 

```{r, warning=FALSE}
# plot 1
plot(data_ygd$y_pred_min, data_ygd$Y, xlim=c(0, 35), ylim=c(0, 35))
abline(a=0, b=1)
abline(a=0, b=0.5)

# plot 2
ggplot(data_ygd, aes(x=Y, y=data_ygd$y_pred_min, ymax = data_ygd$y_pred_min+IYG)) + 
  geom_point(size=2.5, alpha=0.6, color="blue") +
  geom_abline() +
  geom_pointrange(aes(ymin = y_pred_min, ymax = y_pred_min+IYG), data = data_ygd, width = 0.2, size=0.1, linetype='dashed', col = 'red')+
  geom_pointrange(aes(ymin = y_pred_min-UYG, ymax = y_pred_min), data = data_ygd, width = 0.2, size=0.1, linetype='dashed', col = 'green')+
  # Coords and axis limits
  geom_hline(yintercept=max(data_ygd$Y), size=1, linetype='dashed', col = 'red')+
  geom_hline(yintercept=0, linetype='dotted', col = 'red')+
  geom_vline(xintercept=0, linetype='dotted', col = 'red')+
  coord_cartesian(xlim=c(0, max(data_ygd$Y)), ylim=c(0, max(data_ygd$Y))) + 
  labs(title="Yield gap decomposition", subtitle="Following the method of Wairegi et al. (2010)", y="Predicted crop yield Y [t/ha]",
       x="Observed crop yield X [t/ha]", caption="") + theme_bw() + theme(text = element_text(size=15))

```

___

# **Recommendations**

**Please decide how to rephrase these bullets in order to sound less conclusive... but in general these were what I most learnt**

The main advantages of BLA are the following:

•	Empirically based and comprehensible.
•	Lighter computation requirements.
•	Open-source code can be used for implementation (e.g., R-script).
•	Solid web supply of information regarding code functions and libraries to be used.
•	A minimum of two variables are needed (i.e., yield is mandatory plus an additional explaining variable).
•	It does not necessarily require a huge dataset if edaphoclimatic and technological conditions are properly segmented. 


The main disadvantages of BLA are the following:

•	It still requires a relatively large dataset (at least N > 100).
•	The independent variable must cover a wide range of variation, with multiple combinations of Y vs. independent variable within each group of X-variable. 
•	It requires solid control of outliers, mostly on the Y-axis. The error of measured yields must be minimal. Otherwise, the "maximum" approach must be substituted by the "highest_quantile" one, and our experience indicates that this tends to result in a underestimation of the predicted y_max. 
•	The interpretability of the data cannot be purely statistical as the patterns of the 'splines' need to be contextualized through agronomic knowledge. 
•	X-variables with discrete distributions (or X-zero values) must be kept out of this kind of analysis. These may still be used for segmentation of data but its applicability depends largely on the total amount of data and the size of each generated subset of data.
•	Reclassification or segmentation of data according to technological characteristics of the production system (e.g., irrigated vs. rainfed) must be conducted. This implies that even larger datasets are needed when dealing with heterogeneous data in terms of geographic locations, climatic variation, cultivars, or agronomy. Theoretically, this rule would apply to all x-variables which are not entirely independent. However, one majour point of concern is the enormous difficulty that is to identify which variables play an absolutely independent role in crops performance. This is far from being an easy task and that is why we highlight the importance of mechanistic knowledge on crop science to guide the use of this method and the interpretability of results.  
•	Our experience indicates that dataset size must be comprehended between hundreds of observations, when dealing with homogeneous data in terms of agronomy and climate, but this threshold can increase easily to more than a few thousand of observations when climate and agronomy vary. 


___

# **Acknowledgments**

We thank Ken Giller (WUR-PPS) and Piet van Asten (Olam International) for sharing the original data showcased in this workflow. The development of this notebook was possible thanks to the financial support from the OneCGIAR initiative on *Excellence in Agronomy*. For further support and questions on how to implement this workflow to other data sets, or further thoughts on how to improve the theoretical framework used, are almost welcome and should be addressed to **j.silva@cgiar.org**.   

___

# **References**

Casanova, D., Goudriaan, J., Bouma, J., Epema, G.F., 1999. Yield gap analysis in relation to soil properties in direct-seeded flooded rice. Geoderma 91, 191–216.

Casanova, D., Goudriaan, J., Forner, M.M.C., Withagen, J.C.M., 2002. Rice yield prediction from yield components and limiting factors. Eur. J. Agron. 17, 41–61.

French, R.J., Schultz, J.E., 1984. Water use efficiency of wheat in a Mediterranean-type environment. I. The relation between yield, water use and climate. Aust. J. Agric. Res. 35, 743–764.

Fermont, A.M., van Asten, P.J.A., Tittonell, P., van Wijk, M.T., Giller, K.E., 2009. Closing the cassava yield gap: An analysis from smallholder farms in East Africa. Field Crops Res. 112, 24–36.

Grassini, P., Torrion, J.A., Yang, H.S., Rees, J., Andersen, D., Cassman, K.G., Specht, J.E., 2015. Soybean yield gaps and water productivity in the western U.S. Corn Belt. Field Crops Res. 179, 150–163.

Passioura, J.B., Angus, J.F., 2010. Chapter 2 - Improving Productivity of Crops in Water-Limited Environments, in: Sparks, D.L. (Ed.), Advances in Agronomy. Academic Press, pp. 37–75.

Silva, J.V., Reidsma, P., Laborte, A.G., van Ittersum, M.K., 2017. Explaining rice yields and yield gaps in Central Luzon, Philippines: An application of stochastic frontier analysis and crop modelling. Eur. J. Agron. 82, 223–241.

Shatar, T.M., Mcbratney, A.B., 2004. Boundary-line analysis of field-scale yield response to soil properties. J. Agric. Sci. 142, 553–560.

Wairegi, L.W.I., van Asten, P.J.A., Tenywa, M.M., Bekunda, M.A., 2010. Abiotic constraints override biotic constraints in East African highland banana systems. Field Crops Res. 117, 146–153.

Webb, R.A., 1972. Use of the Boundary Line in the analysis of biological data. J. Hortic. Sci. 47, 309–319.  



